Evaluation of dynamic resource allocation strategies for overlay Apache Spark clusters on HPC \tristan{``Overlay'' and ``Apache Spark'' are redundant}

### 2019-05-06 paper.tex
Comments and notes on Section II

    The application, templates, configuration files, benchmarks and analysis
    scripts are publicly available and can be found at:
    \href{https://github.com/big-data-lab-team/spa}{https://github.com/big-data-lab-team/spa}.
    Links to the processing engines and processed data are provided in the text.

    \subsection{Infrastructure}
    All experiments were conducted on the Cedar and B\'eluga HPC computing clusters
    made available by Compute Canada. The Cedar cluster has a total of 1542 nodes
    with a total of 58,416 CPU cores. Available memory on a Cedar node can range
    from 125 to 3022~G. Standard nodes are equipped with either 2x Intel E5-2683 v4
    Broadwell @ 2.1~Ghz (32 cores total) or 2x Intel Platinum 8160F Skylake @
    2.1Ghz (48 cores total) CPUs and 
    2 x 480~G SSD. All nodes and temporary storage on Cedar is connected by an 
    Intel OmniPath (version 1) interconnect with 100Gbit/s bandwidth.

    B\'eluga, on the other hand, is a smaller cluster with 872 available nodes. 
    Node memory can range between 92 to 752~G, with the most common node type having 
    186G of available memory. All nodes contain a total of 2 x Intel Gold 6148
    Skylake @ 2.4~Ghz (40 cores/node) CPU and are connected to each other with a
    100~Gb/s Mellanox Infiniband EDR network. Each non-GPU node type contains one 
    480~G SSD.

    \subsection{Templates}

    Two different templates were used for our experiments: the batch submission
    template and the pilot submission template. The batch submission template was
    inspired by the template provided by Compute Canada to launch Spark applications 
    on Slurm \href{https://docs.computecanada.ca/wiki/Apache\_Spark/en}{available here}.
    The template operates as follows: Certain resource requirements are requested by
    the user (e.g. walltime, amount of memory per node, number of CPUs per task, number 
    of nodes and number of tasks per node). Once these resources are allocated, a 
    master is started on one of the requested node resources. Then, after the master
    has successfully started, the workers are started on all nodes. Multiple worker
    instances are started on a single node by setting the \textit{SPARK\_WORKER\_INSTANCES}
    environment variable to the number of tasks per node. Each worker is given as many as 
    many cores as specified by the user in the SLURM resource allocation request.
    After both the masters and the workers have successfully started, the driver is finally
    started. The amount of memory given to each executor corresponds to 95\% of the 
    available memory on the node, to allow for offheap space. 
    The deploy-mode selected for the batch template is client-mode.

    The pilot submission template is similar to that of the batch template, however,
    each pilot will start its own Spark master and worker. However, only one pilot,
    the first pilot to attempt to launch the driver, will do so. The reason for 
    which each pilot starts its own master is to ensure the fault tolerance of the 
    masters. In this sort of configuration, should the active master be killed, one
    of the stand-by masters will takeover and the application will be able to 
    resume. Such a configuration is particularly favourable in pilot scheduling 
    scenarios as node failures may be more frequent do to walltime expiration.
    Additionally, the deploy mode of the driver was selected to be cluster deploy mode.
    This would not only allow the driver to be executed directly on one of the workers,
    but also allow us to make the driver fault-tolerant through the \textit{supervise} 
    mode, which is only available in cluster deploy. As with the masters, it is
    particularly important to have a fault-tolerant driver in pilot-scheduling scenarios due walltime
    expiration. Should pilots be idle for a certain duration, the pilots will terminate
    themselves such as to not hog resources.

@@

    \subsection{Application}
        \begin{algorithm}\caption{Incrementation}\label{alg:incrementation}              
            \begin{algorithmic}[1]                                                       
                \Input                                                                       
                    \Desc{$x$}{a sleep delay in seconds}                                         
                    \Desc{$n$}{a number of iterations}                                           
                    \Desc{$C$}{a set of image chunks}                                            
                \EndInput                                                                    
                \ForEach{$chunk \in C$}                                                      
                    \State read $chunk$                                        
                    \For{$i \in [1, n]$}                                                         
                        \State $chunk\gets chunk+1$                                              
                        \State sleep $x$                                                         
                    \EndFor                                                                      
                    \State write $chunk$                                            
                \EndFor                                                                      
            \end{algorithmic}                                                                
        \end{algorithm}
    To determine the added value of pilot scheduling over batch scheduling of Spark
    applications, we required a Spark application operating on a large dataset with
    an important processing time to emulate a what would be the average requirements
    of a scientific Spark application. For this, we created a synthetic application 
    that would process the Big Brain~\cite{bigBrain}, a 76G 3D histological
    image of a human brain. The algorithm is a chain of map transformations that
    at each transformation increment the voxels of the image by 1 (see Figure~\ref{alg:incrementation}.
    We chose such a synthetic algorithm
    as the focus of our experiments is pilot scheduling and not the application in
    itself. Furthermore, this algorithm enabled us to have control over the task duration
    which was representative of scientific applications. Additionally, it was important
    that the overall application duration did not vary between the different levels of
    parallelism within our experiments. Being able to adjust the task duration based on
    level of parallelism allowed us to achieve this.

    Investigating the fault-tolerance provided by the overlay cluster is
    crucial in determining the suitable of pilot scheduling with overlay clusters.
    Fault-tolerance of the driver is only possible in cluster deploy mode, however,
    when using Spark's standalone scheduler, this mode is not available for Python 
    applications. It is for this reason that our synthetic application is written
    in Scala. Nevertheless, cluster deploy mode is possible for Python application
    using YARN or Mesos schedulers.

@@

    The makespan of an application can be defined as the total duration of the 
    application including any scheduler queueing time. It can be defined by
    the following expression:

    \begin{equation}
        M = \frac{C}{W} \label{eq:mcw}
    \end{equation}
    where:
    \begin{itemize}
        \item $M$ is the makespan of the application
        \item $C$ is the total CPU time of the application
        \item $W$ is the average number of workers
    \end{itemize}

    The average number of workers allows us to obtain a more accurate makespan
    figure as the average number of workers will decrease with increasing 
    queuing times. Should the average number of workers equal the number of workers
    requested, the application was not subjected to any scheduling queueing time.

    Furthermore, should the total CPU time of the application be fixed, the relation
    between batch and pilot jobs can be represented as:

    \begin{equation}
        \frac{M_{batch}}{M_{pilot}} = f(\frac{W_{pilot}}{W_{batch}})\label{eq:makespancomp}
    \end{equation}
    where:
    \begin{itemize}
        \item $M_{batch}$ is the makespan of the batch application
        \item $M_{pilot}$ is the makespan of the pilot application
        \item $W_{pilot}$ is the average number of workers of the pilot application
        \item $W_{batch}$ is the average number of workers of the batch application
    \end{itemize}

    \todo{perhaps expand a bit here}

@@

        \begin{table}                                                                    
\centering                                                                       
\begin{tabular}{c|c|c}                                                             
\rowcolor{headcolor}                                                             
Configuration & RAM (GB) & Tasks \\                               
\hline                                                                           
1 & 112 & 16 \\                                               
2 & 224 & 32 \\                                               
3 & 336 & 48 \\
4 & 448 & 64 \\
\end{tabular}                                                                    
\setlength{\belowcaptionskip}{-10pt}                                             
\caption{Resource configurations}                                                    
\label{table:dedicatednodes}                                                            
\end{table} 
        \todo{execution of a batch cluster vs dynamic (pilot) cluster}
    \subsection{Robust masters}
        \todo{scala standalone vs our pyspark workaround. kill masters in experiments}
    \subsection{Checkpointing}
        \todo{metric for determining how often to checkpoint based on cluster size}
    \subsection{Job arrays}
        \todo{need to kill idle workers. may not want all workers to be running at once.}
    \subsection{Example application}
        \todo{incrementation with varying task durations}
\section{Results}\label{sec:Results}
\section{Discussion}\label{sec:discussion}
\section{Conclusion}\label{sec:conclusion}

>>>>

    The application, templates, configuration files, benchmarks and
    analysis scripts are publicly available and can be found in our Spark
    Pilot-job scheduler for HPC Applications (SPA) software at:
    \href{https://github.com/big-data-lab-team/spa}{https://github.com/big-data-lab-team/spa}.
    Links to the processing engines and processed data are provided in the
    text.

    \subsection{Infrastructure}
    All experiments were conducted on the Cedar and B\'eluga HPC computing clusters
    made available by Compute Canada through WestGrid and Calcul Qu\'ebec \TG{add links}. Both 
    clusters are accessible through the Slurm batch scheduler and Lustre parallel file system.
    The Cedar cluster has a total of 1542 nodes
    with a total of 58,416 CPU cores. Available memory on a Cedar node can range
    from 125 to 3022~G. Standard nodes are equipped with either 2x Intel E5-2683 v4
    Broadwell @ 2.1~Ghz (32 cores total) or 2x Intel Platinum 8160F Skylake @
    2.1Ghz (48 cores total) CPUs and 
    2 x 480~G SSD. All nodes and temporary storage on Cedar are connected by an 
    Intel OmniPath (version 1) with 100Gbit/s bandwidth.

    B\'eluga, on the other hand, is a smaller cluster with 872 available nodes. 
    Node memory can range between 92 to 752~G, with the most common node type having 
    186G. All nodes contain 2 x Intel Gold 6148
    Skylake @ 2.4~Ghz (40 cores/node) CPU and are connected to each other with a
    100~Gb/s Mellanox Infiniband EDR network \TG{expand EDR}. Each non-GPU node type contains one 
    480~G SSD.

    \subsection{Job Templates}

    Two different job templates were used to implement the two main
    conditions compared in our experiments: the batch submission
    template and the pilot submission template. The batch submission template was
    inspired by the template provided by Compute Canada to launch Spark applications 
    on Slurm \href{https://docs.computecanada.ca/wiki/Apache\_Spark/en}{available here}.
    The template operates as follows: Certain resource requirements are requested by
    the user (e.g. walltime, amount of memory per node, number of CPUs per task, number 
    of nodes and number of tasks per node). Once these resources are allocated, a 
    master is started on one of the requested node resources. Then, after the master
    has successfully started, the workers are started on all nodes. Multiple worker
    instances are started on a single node by setting the \textit{SPARK\_WORKER\_INSTANCES}
    environment variable to the number of tasks per node. Each worker is given as many
     cores as specified by the user in the Slurm resource allocation request.
    After both the masters and the workers have successfully started, the driver is finally
    started. The amount of memory given to each executor corresponds to 95\% of the 
    available memory on the node, to allow for offheap space. 
    The Spark deploy-mode selected for the batch template is client-mode.

    The pilot submission template is similar to that of the batch template, however,
    each pilot will start its own Spark master and worker. Only one pilot,
    the first pilot to attempt to launch the driver, will do so. The reason for 
    which each pilot starts its own master is to ensure the fault tolerance of the 
    masters. In this sort of configuration, should the active master be killed, one
    of the stand-by masters will takeover and the application will be able to 
    resume. Such a configuration is particularly favourable in pilot scheduling 
    scenarios as node failures may be more frequent due to walltime expiration.
    Additionally, the Spark deploy mode of the driver was selected to be cluster deploy mode.
    This would not only allow the driver to be executed directly on one of the workers,
    but also allow us to make the driver fault-tolerant through the
    \textit{supervise} mode, which is only available in cluster deploy
    \TG{A brief section on Spark to explain these modes and features would
    be useful}. As with the masters, it is particularly important to have a
    fault-tolerant driver in pilot-scheduling scenarios due to possible
    walltime expiration \TG{It should be explained before that one of the interests of pilots
    is to ``play'' with shorter walltimes}. \TG{It would also be nice to have runs where pilots 
    have shorter walltimes, to see if queuing times are shorter too.}
    Should pilots be idle for a certain duration, the
    pilots will terminate themselves such as to not hog resources.

@@

    \subsection{Application}
        \begin{algorithm}\caption{Incrementation}\label{alg:incrementation}              
            \begin{algorithmic}[1]                                                       
                \Input                                                                       
                    \Desc{$x$}{a sleep delay in seconds}                                         
                    \Desc{$n$}{a number of iterations}                                           
                    \Desc{$C$}{a set of image chunks}                                            
                \EndInput                                                                    
                \ForEach{$chunk \in C$}                                                      
                    \State read $chunk$                                        
                    \For{$i \in [1, n]$}                                                         
                        \State $chunk\gets chunk+1$                                              
                        \State sleep $x$ \TG{Don't you sleep by x minus the duration of the previous step?}                                                        
                    \EndFor                                                                      
                    \State write $chunk$                                            
                \EndFor                                                                      
            \end{algorithmic}                                                                
        \end{algorithm}
    To determine the added value of pilot scheduling over batch scheduling of Spark
    applications, we required a Spark application operating on a large dataset with
    an important processing time to emulate a what would be the average requirements
    of a scientific Spark application. For this, we created a synthetic application 
    that would process the Big Brain~\cite{bigBrain}, a 76~G 3D histological
    image of a human brain. The algorithm is a chain of map transformations that
    at each transformation increment the voxels of the image by 1 (see Algorithm~\ref{alg:incrementation}).
    We chose such a synthetic algorithm
    as the focus of our experiments is pilot scheduling and not the application in
    itself. Furthermore, this algorithm enabled us to have control over the task duration
    which was representative of scientific applications. Additionally, it was important
    that the overall application duration did not vary between the different levels of
    parallelism within our experiments. Being able to adjust the task duration based on
    level of parallelism allowed us to achieve this.

    Investigating the fault-tolerance provided by the overlay cluster is
    crucial in determining the suitability of pilot scheduling with overlay clusters.
    Fault-tolerance of the driver is only possible in cluster deploy mode, however,
    when using Spark's standalone scheduler, this mode is not available for Python 
    applications. It is for this reason that our synthetic application is written
    in Scala. Nevertheless, cluster deploy mode is possible for Python application
    using YARN or Mesos schedulers.

@@

    The makespan of an application can be defined as the total duration
    between the submission time of the first application task, and the
    completion time of the last application time. It includes any
    scheduling time, queueing time, data transfer time, and any other
    overhead.

    Assuming a divisible load, i.e., the application can be divided in any
    number of tasks, the makespan can be expressed using the following
    expression, which holds for both batch and pilot execution modes:
    \begin{equation}
        M = \frac{C}{W} \label{eq:mcw}
    \end{equation}
    where:
    \begin{itemize}
        \item $M$ is the makespan of the application
        \item $C$ is the total CPU time of the application
        \item $W$ is the average number of Spark workers throughout the execution
    \end{itemize}
    The average number of workers $W$ allows us to obtain a more accurate makespan
    figure as the average number of workers will decrease with increasing 
    queuing times. It is computed as follows:
    $$
    W = \frac{1}{M}\int_0^M{w(t)dt},
    $$
    where $w(t)$ is the number of workers available at time $t$. When the
    application is not subject to any scheduling or queuing time, the
    average number of workers equal the number of workers requested.

    Therefore, assuming a fixed total CPU time, the relation
    between batch and pilot jobs can be represented as:
    \begin{equation}
        \frac{M_{batch}}{M_{pilot}} = \frac{W_{pilot}}{W_{batch}}\label{eq:makespancomp}
    \end{equation}
    where:
    \begin{itemize}
        \item $M_{batch}$ is the makespan of the batch application
        \item $M_{pilot}$ is the makespan of the pilot application
        \item $W_{pilot}$ is the average number of workers of the pilot application
        \item $W_{batch}$ is the average number of workers of the batch application
    \end{itemize}
    We will use this relation to discuss our results later on. It
    corresponds to an ideal case where no data or other overhead is
    present: only queuing times are included.

    \todo{perhaps expand a bit here \TG{I have a formal proof for eq (1) but I don't think
    it's required here.}}

@@

        \begin{table}                                                                    
\centering                                                                       
\begin{tabular}{c|c|c}                                                             
\rowcolor{headcolor}                                                             
Configuration & RAM (GB) & Tasks \\                               
\hline                                                                           
1 & 112 & 16 \\                                               
2 & 224 & 32 \\                                               
3 & 336 & 48 \\
4 & 448 & 64 \\
\end{tabular}                                                                    
\setlength{\belowcaptionskip}{-10pt}                                             
\caption{Resource configurations}                                                    
\label{table:dedicatednodes}                                                            
\end{table} 
        \todo{execution of a batch cluster vs dynamic (pilot) cluster}
    \subsection{Robust masters}
        \todo{scala standalone vs our pyspark workaround. kill masters in experiments}
    \subsection{Checkpointing}
        \todo{metric for determining how often to checkpoint based on cluster size}
    \subsection{Job arrays}
        \todo{need to kill idle workers. may not want all workers to be running at once.}
    \subsection{Example application}
        \todo{incrementation with varying task durations}
\section{Results}\label{sec:Results}
\section{Discussion}\label{sec:discussion}

SPA vs other frameworks.

Technical overhead of using pilots + an overlay cluster over a cluster.

\section{Conclusion}\label{sec:conclusion}
